---
title: "Data Exploration and Preparation"
author: "Marco Scutari"
date: "`r Sys.Date()`"
output: html_document
---

```{r load_libraries, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE, results = FALSE}
library(knitr)
library(bnlearn)
library(lattice)
library(latticeExtra)
library(gridExtra)

knitr::opts_chunk$set(fig.align = "center")

standard.barplot = function(x, xlab = "number of observations") {

  my.padding = list(
      layout.heights = list(
        top.padding = 0,
        main.key.padding = 0,
        key.axis.padding = 0,
        axis.xlab.padding = 0,
        xlab.key.padding = 0,
        key.sub.padding = 0),
      layout.widths = list(
        left.padding = 0,
        key.ylab.padding = 0,
        ylab.axis.padding = 0,
        axis.key.padding = 0,
        right.padding = 0)
      )

  barchart(rev(x), horizontal = TRUE, col = "skyblue",
    xlim = c(0, 1.05), xlab = xlab, 
    panel = function(...) {
      
      panel.barchart(...)
      panel.grid(h = 0, v = -1)
      
    }, par.settings = my.padding)

}#STANDARD.BARPLOT

standard.heatmap = function(mat, numbers = TRUE, digits = 2, cex = 0.8, 
    abs = FALSE, main = "") {

  col.l = colorRampPalette(c('ivory', 'tomato'))(30)

  my.padding = list(
      layout.heights = list(
        top.padding = 0,
        main.key.padding = 0,
        key.axis.padding = 0,
        axis.xlab.padding = 0,
        xlab.key.padding = 0,
        key.sub.padding = 0),
      layout.widths = list(
        left.padding = 0,
        key.ylab.padding = 0,
        ylab.axis.padding = 0,
        axis.key.padding = 0,
        right.padding = 0),
      panel.background = list(
        col = "grey90"
       )
      )

  if (abs)
    transformed = abs(mat)
  else
    transformed = mat

  levelplot(transformed, col.regions = col.l, colorkey = FALSE,
    scales = list(x = list(rot = 90)), xlab = "", ylab = "", main = main,
    panel = function(y, x, z, ...) {

      panel.levelplot(y = y, x = x, z = z, ...)

      if (numbers) {

        if (abs)
          rounded = as.character(round(mat[cbind(x, y)], digits))
        else
          rounded = as.character(round(z, digits))
        rounded[is.na(rounded)] = ""
        ltext(x = x, y = y, labels = rounded, cex = cex)

      }#THEN

    }, par.settings = my.padding
  )

}#STANDARD.HEATMAP

standard.histogram = function(var, legend) {

  histogram(var,
    scales = list(tck = c(1, 0)),
    xlab = legend, ylab = "percent of total", col = "skyblue",
    panel = function(...) {

      panel.grid(h = -1, v = 0)
      panel.histogram(...)

  })

}#STANDARD.HISTOGRAM

standard.xyplot = function(formula, xlab, ylab, regression = FALSE) {

  xyplot(formula, data = data, pch = 19, 
    scales = list(tck = c(1, 0)),
    xlab = xlab, ylab = ylab,
    panel = function(...) {

      panel.grid(h = -1, v = -1)
      panel.xyplot(..., col = "skyblue")
      if (regression)
        panel.smoother(..., col = "skyblue3", lwd = 3, span = 1/2, level = 0.95)

  })

}#STANDARD.XYPLOT

prepare.map = function(data, fun) {

  latitudes = sort(unique(data[, "lat"]))
  longitudes = sort(unique(data[, "lon"]), decreasing = FALSE)
  map = matrix(0, nrow = length(longitudes), ncol = length(latitudes), 
            dimnames = list(longitude = longitudes, latitude = latitudes))

  for (i in seq_along(longitudes))
    for (j in seq_along(latitudes)) {

      location = subset(data, (lon == longitudes[i]) & (lat == latitudes[j]))

      if (nrow(location) == 0)
        map[i, j] = NA
      else
        map[i, j] = fun(location)

    }#FOR

  map[is.nan(map)] = NA

  return(map)

}#PREPARE.MAP

species = c("herring", "sprat", "cod", "flounder", "whitefish", "salmon", 
            "trout", "smelt", "bream", "ide", "roach", "pike", "perch",
            "sander", "burbot", "vendace")
```

# Load and Subset the Data

```{r load_the_data}
data = readRDS("../data/prepd.rds")
dim(data)
```

* Remove areas far from the Finnish coast: the catch data comes from Finnish 
  professional fishers, so the data is not representative of the total catches 
  of the far-away areas.

```{r restrict_to_close_coastal_areas}
close.to.finland = 
  c("2", "3", "5", "6", "7", "10", "11", "12", "14", "15", "16", "19", "20", 
    "22", "23", "24", "27", "28", "31", "32", "36", "37", "41", "42", "45",
    "46", "47", "49", "50", "51", "52", "53", "54", "55", "56", "57", "59", 
    "60", "61", "62", "63", "64")

data = data[data$rectangle %in% close.to.finland, ]
```

* Observations with missing months are self-reporting over the whole year. We
  could assume that the catches have the same distribution over the months as
  the rest of the data, but we just drop them for simplicity.

```{r remove_unknown_months}
data = data[!is.na(data$month), ]
```

```{r factor_cleanup, echo = FALSE}
for (var in names(data))
  if (is.factor(data[, var]))
    data[, var] = droplevels(data[, var])
```

# Explore the Missing Values

The proportion of missing values is negligible (< 10%) for all the geographical
and catch variables. It is significant (> 30%) for all the variables that
measure the water characteristics and for `surface`.

```{r proportions_of_missing_values, echo = FALSE, fig.width = 8, fig.height = 8}
local({

  props = sapply(data, function(x) mean(is.na(x)))

  standard.barplot(props, xlab = "proportion of missing data")

})
```

The distribution of missing values over the years is approximately constant,
suggesting that the pattern of missingness is not time-dependent. Or at least
that it does not have a trend.

```{r missing_values_over_years, echo = FALSE, fig.width = 8, fig.height = 8}
local({

  props = numeric(nlevels(data$year))
  names(props) = levels(data$year)

  for (current in names(props)) {

    period = subset(data, year == current)
    props[current] = sum(sapply(period, function(x) sum(is.na(x)))) /
                       (nrow(period) * ncol(period))

  }#FOR

  standard.barplot(props, xlab = "proportion of missing data")

})
```

The missing values do not appear to have a seasonal component; their proportion
remains relatively constant over the year. There seems to be a shallow peak in
winter and around New Year.

```{r missing_values_over_months, echo = FALSE, fig.width = 8, fig.height = 4}
local({

  props = numeric(nlevels(data$month))
  names(props) = levels(data$month)

  for (current in names(props)) {

    period = subset(data, month == current)
    props[current] = sum(sapply(period, function(x) sum(is.na(x)))) /
                       (nrow(period) * ncol(period))

  }#FOR

  standard.barplot(props, xlab = "proportion of missing data")

})
```

What is the geographical distribution of the missing data?

```{r missing_data_over_space, echo = FALSE, fig.width = 8, fig.height = 8}
local({

  props = prepare.map(data, fun = function(loc) {

    sum(sapply(loc, function(x) sum(is.na(x)))) / (nrow(loc) * ncol(loc))

  })

  standard.heatmap(props)

})
```

There are just three locations with proportions of missing data greater than 
30%. The reason is that there are (years, months) in which no species catches
are measured at all. For the rest, all the species are monitored. It happens in
(59.75, 25.5):

```{r location_with_many_missing_1, echo = FALSE}
local({

  location = subset(data, (lat == 59.75) & (lon == 25.5))
  props = apply(location[, species], 1, function(x) sum(is.na(x)))
  props = paste(props, "/", length(species))
  kable(table(props))

})
```

It happens in (65.25, 23.5):

```{r location_with_many_missing_2, echo = FALSE}
local({

  location = subset(data, (lat == 65.25) & (lon == 23.5))
  props = apply(location[, species], 1, function(x) sum(is.na(x)))
  props = paste(props, "/", length(species))
  kable(table(props))

})
```

And it happens (65.25, 23.5):

```{r location_with_many_missing_3, echo = FALSE}
local({

  location = subset(data, (lat == 64.25) & (lon == 22.5))
  props = apply(location[, species], 1, function(x) sum(is.na(x)))
  props = paste(props, "/", length(species))
  kable(table(props))

})
```

# Clean Up the Missing Values

Based on the findings above, we:

* Remove the `chlorophyll.a`, `suspended.solids` and `ice` variables because
  more than 60% of their values are missing.

```{r remove_variables}
data = data[, setdiff(colnames(data), c("chlorophyll.a", "suspended.solids", "ice"))]
```

* Remove the observations for which all species catches are missing values.

```{r remove_observations_with_no_catches}
missing.species = apply(data[, species], 1, function(x) sum(is.na(x)))
data = data[missing.species != length(species), ]
```

```{r define_time_periods, echo = FALSE}
decades = ordered(gsub("[0-9]$", "0", data[, "year"]))
seasons = factor(data[, "month"], labels = c(rep("winter", 2), 
            rep("spring", 3), rep("summer", 3), rep("autumn", 3), "winter"))
```

# Explore the Catch Variables

The yearly volume of catch across all species and locations has a cyclic pattern
with a period of about 15 years. Here plotted with 95% confidence bands.

```{r total_volume_of_catches, echo = FALSE, fig.width = 9, fig.height = 4}
local({
  
  df = aggregate(data[, species], by = data[, "year", drop = FALSE], FUN = sum)
  df = data.frame(year = df[, "year"], catch = rowSums(df[, -1]))

  xyplot(catch ~ year, data = df, col = "skyblue", 
    ylab = "total catch", xlab = "year",
    scales = list(tck = c(1, 0), alternating = 1,
                  x = list(at = c(1 + 0:4 * 10), rot = 45)),
    panel = function(x, y, ...) {

      panel.abline(h = mean(y), col = "lightgrey")
      panel.smoother(y = y, x = x, ..., lwd = 2, span = 1/3, level = 0.95)

    })

})
```

The catches for each species over all locations evolve over time. Here they are
standardised and plotted with 95% confidence bands.

```{r overall_volume_of_catches_per_species, echo = FALSE, fig.width = 10, fig.height = 8}
local({

  overall.catch = sapply(species, function(s) {
    aggregate(data[, s], by = data[, "year", drop = FALSE], FUN = sum)
  }, simplify = FALSE)

  df = sapply(overall.catch, function(species) scale(species[, "x"]))
  df = data.frame(year = overall.catch[[1]][, 1], df)
  df = reshape2::melt(df, id.var = "year", value.name = "catch", variable.name = "species")

  print(xyplot(catch ~ year | species, data = df, col = "skyblue",
    ylab = "standardized catch", xlab = "year",
    scales = list(tck = c(1, 0), alternating = 1, 
                  x = list(at = c(1 + 0:4 * 10), rot = 45)), 
    panel = function(...) {

      panel.abline(h = 0, col = "lightgrey")
      panel.smoother(..., lwd = 2, span = 1/3, level = 0.95)

    }))

})
```

The catches overall have a seasonality that comes from fishing seasons: they
concentrate between April and July.

```{r seasonality_of_aggregated_catches, echo = FALSE, fig.width = 9, fig.height = 4}
local({

  df = aggregate(data[, species], by = data[, "month", drop = FALSE], FUN = sum)
  df = data.frame(month = df[, "month"], catch = rowSums(df[, -1]))

  xyplot(catch ~ month, data = df, col = "skyblue",
    ylab = "total catch", xlab = "month",
    scales = list(tck = c(1, 0), alternating = 1),
    panel = function(x, y, ...) {

      panel.abline(h = mean(y), col = "lightgrey")
      panel.abline(v = c(4, 7), col = "lightgrey", lty = 2)
      panel.smoother(y = y, x = x, ..., lwd = 2, span = 1/3, level = 0.95)

    })

})
```

As expected, most species are caught for the most part between April and July.
A few exceptions, like sprat, vendace, burbot, are caught primarily
in winter. They may account for the slight increase in catch between November
and January.

```{r seasonality_of_catches_by_species, echo = FALSE, fig.width = 10, fig.height = 8}
local({

  overall.catch = sapply(species, function(s) {
    aggregate(data[, s], by = data[, "month", drop = FALSE], FUN = sum)
  }, simplify = FALSE)

  df = sapply(overall.catch, function(species) scale(species[, "x"]))
  df = data.frame(month = overall.catch[[1]][, 1], df)
  df = reshape2::melt(df, id.var = "month", value.name = "catch", variable.name = "species")

  print(xyplot(catch ~ month | species, data = df, col = "skyblue",
    ylab = "standardized catch", xlab = "month",
    scales = list(tck = c(1, 0), alternating = 1),
    panel = function(...) {

      panel.abline(v = c(4, 7), col = "lightgrey", lty = 2)
      panel.abline(h = 0, col = "lightgrey")
      panel.smoother(..., lwd = 2, span = 1/3, level = 0.95)

    }))

})
```

The Spearman correlation between species is mostly positive, which suggests that
species are not competing for the same (limited) resources in the same areas.

> **Laura:** There are two mechanisms that tie some of the species together: The
> ones that are most correlated (bream, roach, ide, pike, perch) both line in
> rather similar habitats (coastal, shallow, near islands, etc.), and they most
> likely also come from the same fishery: coastal gill-nets and other
> non-discriminating gear. On the other hand, this should also mean that the
> fish come in approximately the same proportions to the catch than they are in
> the nature (although of course how actively they move etc. is species-specific
> to some extent, which biases this ratio). 
> 
> The only quota species here are salmon, herring, and sprat, the latter of
> which has not traditionally been a targeted species in Finland. There are some
> technical restrictions in the fishery of other species too, such as minimum
> landing sizes, but I would not expect them to drive the catches. 

```{r overall_correlation_between_species, echo = FALSE, fig.width = 8, fig.height = 8}
local({

  cormat = cor(data[, species], method = "spearman")
  standard.heatmap(cormat, abs = TRUE)

})
```

These correlations appear to change over time, suggesting that the (vector time
series) of the catches might not be homogeneous. The most evident change is the
increasing correlation between sprat and herring in the bottom left corner. The
correlation between whitefish and several other fishes appears to increase with
time.

```{r correlations_between_species_over_time, echo = FALSE, fig.width = 8, fig.height = 12}
local({
  
  figures = vector(nlevels(decades), mode = "list")
  names(figures) = levels(decades)

  for (d in levels(decades)) {

    cormat = cor(data[decades == d, species], method = "spearman")
    figures[[d]] = standard.heatmap(cormat, numbers = FALSE, abs = TRUE, main = d)

  }#FOR

  do.call(grid.arrange, c(figures, list(ncol = 2)))

})
```

The spatial distribution of the catches varies between different species: there
are a few locations in which many species are caught (ide, roach, pike, perch,
burbot in (60.75, 2.15) and surrounding locations), but there also are species
that a caught in completely different places (vendace). In addition, there are
species that are caught in just a single location (vendace, ide, roach, pike,
perch, burbot) and species that are caught all over the monitored area (trout,
salmon).

```{r species_over_space, echo = FALSE, fig.width = 12, fig.height = 16}
local({

  figures = vector(length(species), mode = "list")
  names(figures) = species

  for (s in species) {

    props = prepare.map(data, fun = function(loc) sum(loc[, s]))
    figures[[s]] = standard.heatmap(props, numbers = FALSE, main = s)

  }#FOR

  do.call(grid.arrange, c(figures, ncol = 4))

})
```

> **Laura:** Basically, I would expect certain factors, such as temperature or
> its proxies, to explain a lot of variability in the catches of some species,
> while the geography may be more important for others. If we can find evidence
> of the actual water quality variables having an effect, that would be nice.

Below are the temperatures recorded for each species, that is, the temperatures
in those observations in which the species catch is strictly positive. Vendace
has a profile that is clearly different from the other species. Other
differences are more subtle. Some species have a second, smaller peak just above
15C, while others do not.

```{r species_temperature_profiles, echo = FALSE, fig.width = 12, fig.height = 24}
local({

  df = lapply(species, function(s) {

    record = !is.na(data[, s]) & !is.na(data[, "temperature"]) & (data[, s] > 0)
    temp = data[record, "temperature"]
    data.frame(species = rep(s, length(temp)), temperature = temp)

  })

  df = do.call(rbind, df)
  df$species = factor(df$species, levels = species)

  histogram(~ temperature | species, data = df, layout = c(1, length(species)),
    scales = list(tck = c(1, 0), alternating = 1),
    xlim = c(-10, 30), strip = FALSE,
    panel = function(...) {

      panel.grid(h = -1, v = 0, col = "lightgrey")
      panel.text(x = 28, y = 17, labels = species[panel.number()], pos = 2)
      panel.histogram(...)
  
    })

})
```

If we tally the catches across temperature ranges, we can see some more
differences among species. The plots below show the frequency of temperatures
(*x*-axis) and normalised caches (*y*-axis) jointly-discretised values. Note
that both variables, so the intervals differ between plots, should only be
interpreted in relative terms.

For some species, there is a strong association between temperature and catches.
Burbot and pike are only caught in large quantities at low temperatures. Perch
catches appear to increase with temperature. On the other hand, salmon is only
caught in large quantities at higher temperatures. Pike is mainly caught in
large quantities at lower temperatures.

```{r catches_temperature_profiles, echo = FALSE, fig.width = 18, fig.height = 18}
local({

  df = lapply(species, function(s) {

    record = !is.na(data[, s]) & !is.na(data[, "temperature"]) & (data[, s] > 0)
    temp = data[record, "temperature"]
    catch = data[record, s]
    catch = (catch - min(catch)) / (max(catch) - min(catch))
    data.frame(species = rep(s, length(temp)), 
               temperature = temp,
               catch = catch)

  })

  df = do.call(rbind, df)
  df$species = factor(df$species, levels = species)

  figures = vector(nlevels(species), mode = "list")
  names(figures) = levels(species)

  for (s in species) {

    tabulated = discretize(df[df$species == s, ], method = "hartemink",
                  breaks = c(2, 5, 7), ibreaks = c(5, 200, 200))
    tabulated = table(tabulated)


    figures[[s]] = standard.heatmap(tabulated[s, , ], main = s)

  }#FOR

  do.call(grid.arrange, c(figures, list(ncol = 4)))

})
```

# Explore the Other Variables

The data contain several variables describing the characteristics of each
location. In addition to the coordinates of its bounding box (`lon.min`,
`lon.max`, `lat.min`, `lat.max`), we have the actual sea `water.area` and how
much of it is `open.water` and `coastal.zone`.

The actual sea area is smaller than that in the bounding box and varies
in [`r range(data[, "water.area"], na.rm = TRUE)`] depending on the coastline.

```{r histogram_of_areas, echo = FALSE}
standard.histogram(data[, "water.area"], legend = "sea area")
```

The total area within the bounding box in each location varies in
[`r range(data[, "area"], na.rm = TRUE)`], so the proportion of water area is
distributed as follows.

```{r histogram_of_water_area_proportions, echo = FALSE}
standard.histogram(data[, "water.area"] / data[, "area"], legend = "sea area")
```

All the missing values for this variable are associated with a single location.

```{r location_with_unknown_area}
unique(data[is.na(data[, "area"]), c("lon", "lat")])
unique(data[is.na(data[, "water.area"]), c("lon", "lat")])
```

---

The proportion of open sea varies by location, from almost none to completely
open sea. As expected, it is greater in locations farther from the coast.
The proportion of coastal area is greater for locations close to the coasts and
is the complement to the open sea areas.

```{r open_sea_by_location, echo = FALSE, fig.width = 12, fig.height = 8}
local({

  props = prepare.map(data, 
            fun = function(loc) mean(loc[, "open.water"] / loc[, "water.area"]))
  p1 = standard.heatmap(props, main = "open water")
  props = prepare.map(data, 
            fun = function(loc) mean(loc[, "open.water"] / loc[, "water.area"]))
  p2 = standard.heatmap(props, main = "coastal zone")

  grid.arrange(p1, p2, ncol = 2)

})
```

The missing values for both variables are associated with the same location for
which we do not observe `area` and `water.area`.

```{r location_with_unknown_water_areas}
unique(data[is.na(data[, "open.water"]), c("lon", "lat")])
unique(data[is.na(data[, "coastal.zone"]), c("lon", "lat")])
```

---

The surface sea temperatures range from freezing to just cool. They vary over
seasons and locations, sometimes dramatically and sometimes not. Missing values
can be found in all locations and all years/months.

```{r temperature_by_location, echo = FALSE, fig.width = 8, fig.height = 12}
local({

  figures = vector(nlevels(seasons), mode = "list")
  names(figures) = levels(seasons)

  for (s in levels(seasons)) {

    props = prepare.map(data[seasons == s, ],
              fun = function(loc) mean(loc[, "temperature"], na.rm = TRUE))
    figures[[s]] = standard.heatmap(props, main = s, cex = 0.6)

  }#FOR

  do.call(grid.arrange, c(figures, ncol = 2))

})
```

The distribution of the temperatures does not appear to change over time.

```{r temperature_over_time, echo = FALSE, fig.width = 16, fig.height = 5}
local({
  
  figures = vector(nlevels(decades), mode = "list")
  names(figures) = levels(decades)

  for (d in levels(decades)) {

    props = prepare.map(data[decades == d, ],
             fun = function(loc) mean(loc[, "temperature"], na.rm = TRUE))
    figures[[d]] = standard.heatmap(props, numbers = FALSE, main = d)

  }#FOR

  do.call(grid.arrange, c(figures, list(ncol = 5)))

})
```

---

The amount of phosphorus and nitrogens are forms of pollution (from agriculture?)
and are related as a result.

```{r compare_phosphorus_and_nitrogens, echo = FALSE}
standard.xyplot(log(nitrogen) ~ log(phosphorus), regression = TRUE,
  xlab = "phosphorus (log-scale)", ylab = "nitrogens (log-scale)")
```


Therefore, they have a similar distribution over locations. With over 40% missing
values, there are no complete data for any locations or years.

```{r pollution_by_location, echo = FALSE, fig.width = 12, fig.height = 8}
local({

  latitudes = sort(unique(data$lat))
  longitudes = sort(unique(data$lon), decreasing = FALSE)

  props = prepare.map(data, 
            fun = function(loc) mean(loc[, "phosphorus"], na.rm = TRUE)) 
  p1 = standard.heatmap(props, main = "phosphorus")
  props = prepare.map(data, 
            fun = function(loc) mean(loc[, "nitrogen"], na.rm = TRUE)) 
  p2 = standard.heatmap(props, main = "nitrogen")

  grid.arrange(p1, p2, ncol = 2)

})
```

The distribution of `nitrogen` over locations does not appear to change over
time.

```{r nitrogen_over_time, echo = FALSE, fig.width = 16, fig.height = 5}
local({
  
  figures = vector(nlevels(decades), mode = "list")
  names(figures) = levels(decades)

  for (d in levels(decades)) {

    props = prepare.map(data[decades == d, ],
              fun = function(loc) mean(loc[, "nitrogen"], na.rm = TRUE))
    figures[[d]] = standard.heatmap(props, numbers = FALSE, main = d)

  }#FOR

  do.call(grid.arrange, c(figures, list(ncol = 5)))

})
```

The distribution of `phosphorus` *does* appear to change in the 2020s.

```{r phosphorus_over_time, echo = FALSE, fig.width = 16, fig.height = 5}
local({
  
  figures = vector(nlevels(decades), mode = "list")
  names(figures) = levels(decades)

  for (d in levels(decades)) {

    props = prepare.map(data[decades == d, ],
              fun = function(loc) mean(loc[, "phosphorus"], na.rm = TRUE))
    figures[[d]] = standard.heatmap(props, numbers = FALSE, main = d)

  }#FOR

  do.call(grid.arrange, c(figures, list(ncol = 5)))

})
```

Nitrogens present systematic seasonal patterns: in each location, it is higher
in winter/spring than in summer/autumn.

```{r nitrogen_by_season, echo = FALSE, fig.width = 8, fig.height = 12}
local({

  figures = vector(nlevels(seasons), mode = "list")
  names(figures) = levels(seasons)

  for (s in levels(seasons)) {

    props = prepare.map(data[seasons == s, ],
              fun = function(loc) mean(loc[, "nitrogen"], na.rm = TRUE))
    figures[[s]] = standard.heatmap(props, main = s, cex = 0.6, digits = 0)

  }#FOR

  do.call(grid.arrange, c(figures, ncol = 2))

})
```

A similar pattern appears to hold for `phosphorus`.

```{r phosphorus_by_season, echo = FALSE, fig.width = 8, fig.height = 12}
local({

  figures = vector(nlevels(seasons), mode = "list")
  names(figures) = levels(seasons)

  for (s in levels(seasons)) {

    props = prepare.map(data[seasons == s, ],
              fun = function(loc) mean(loc[, "phosphorus"], na.rm = TRUE))
    figures[[s]] = standard.heatmap(props, main = s, cex = 0.6, digits = 0)

  }#FOR

  do.call(grid.arrange, c(figures, ncol = 2))

})
```

---

The variables `turbidity` and `secchi` both measure how clear the water is,
albeit on opposite scales.

> **Laura:** Since they have about the same amount of missing values, I suggest
> we can drop Secchi and keep turbidity, out of the two.

```{r compare_turbidity_and_secchi, echo = FALSE}
standard.xyplot(log(turbidity) ~ I(-log(secchi)), regression = TRUE,
  xlab = "- secchi (log-scale)", ylab = "turbidity (log-scale)")
```

This relationship is broadly reflected in the values in all locations.

```{r clear_water_by_location, echo = FALSE, fig.width = 12, fig.height = 8}
local({

  props = prepare.map(data,
            fun = function(loc) mean(loc[, "turbidity"], na.rm = TRUE))
  p1 = standard.heatmap(props, main = "turbidity")
  props = prepare.map(data,
            fun = function(loc) mean(loc[, "secchi"], na.rm = TRUE))
  p2 = standard.heatmap(props, main = "secchi")

  grid.arrange(p1, p2, ncol = 2)

})
```

The distribution of `turbidity` over locations does not appear to change over
time.

```{r turbidity_over_time, echo = FALSE, fig.width = 16, fig.height = 5}
local({
  
  figures = vector(nlevels(decades), mode = "list")
  names(figures) = levels(decades)

  for (d in levels(decades)) {

    props = prepare.map(data[decades == d, ],
              fun = function(loc) mean(loc[, "turbidity"], na.rm = TRUE))
    figures[[d]] = standard.heatmap(props, numbers = FALSE, main = d)

  }#FOR

  do.call(grid.arrange, c(figures, list(ncol = 5)))

})
```

Neither does that of `secchi`.

```{r secchi_over_time, echo = FALSE, fig.width = 16, fig.height = 5}
local({
  
  figures = vector(nlevels(decades), mode = "list")
  names(figures) = levels(decades)

  for (d in levels(decades)) {

    props = prepare.map(data[decades == d, ],
              fun = function(loc) mean(loc[, "secchi"], na.rm = TRUE))
    figures[[d]] = standard.heatmap(props, numbers = FALSE, main = d)

  }#FOR

  do.call(grid.arrange, c(figures, list(ncol = 5)))

})
```

Turbidity does not present any particular seasonal patterns.

```{r turbidity_by_season, echo = FALSE, fig.width = 8, fig.height = 12}
local({

  figures = vector(nlevels(seasons), mode = "list")
  names(figures) = levels(seasons)

  for (s in levels(seasons)) {

    props = prepare.map(data[seasons == s, ],
              fun = function(loc) mean(loc[, "turbidity"], na.rm = TRUE))
    figures[[s]] = standard.heatmap(props, main = s, cex = 0.6)

  }#FOR

  do.call(grid.arrange, c(figures, ncol = 2))

})
```

Neither does `secchi`.

```{r secchi_by_season, echo = FALSE, fig.width = 8, fig.height = 12}
local({

  figures = vector(nlevels(seasons), mode = "list")
  names(figures) = levels(seasons)

  for (s in levels(seasons)) {

    props = prepare.map(data[seasons == s, ],
              fun = function(loc) mean(loc[, "secchi"], na.rm = TRUE))
    figures[[s]] = standard.heatmap(props, main = s, cex = 0.6)

  }#FOR

  do.call(grid.arrange, c(figures, ncol = 2))

})
```

> **Laura:** Secchi would be expected to depend on both turbidity and water
> colour.

The distribution of the colour number is bell-shaped on a log-scale but somewhat
asymmetric, with a short left tail. The reason is that water colour is defined
to be positive on the natural scale.

```{r water_colour, echo = FALSE}
local({

  standard.histogram(log(data$colour.number), legend = "colour number")

})
```

> **Laura:** The colour value is an old technique, traditionally done visually
> by comparing the water to standardised colour samples.
>
> https://assets.thermofisher.com/TFS-Assets/LPD/Application-Notes/an_034_tip_color_measurement_1120.pdf
>
> It's usually related to humic components that are dissolved in the water and
> make the water appear brown – in contrast to turbidity, which consists of
> non-dissolved components (often clay etc.) that scatter light. There is no
> theoretical upper limit to colour value as far as I'm aware. Waters from bog
> ponds can reach values up to 200 or so.

Broadly, `colour.number` takes lower values away from the coast. It gets higher
in the northern locations and is exceptionally high only in one location.
This pattern becomes more apparent if we count the proportion of observations
which take `colour.number` values greater than 25 at each location.

```{r colour_number_by_location, echo = FALSE, fig.width = 8, fig.height = 6}
local({

  props = prepare.map(data,
            fun = function(loc) mean(loc[, "colour.number"], na.rm = TRUE))
  p1 = standard.heatmap(props, digits = 1, cex = 0.7)

  props = prepare.map(data,
            fun = function(loc) mean(loc[, "colour.number"] > 25, na.rm = TRUE))
  p2 = standard.heatmap(props, digits = 1, cex = 0.7)

  grid.arrange(p1, p2, ncol = 2)

})
```

Secchi is marginally correlated with both `turbidity` and `colour.number`, is
correlated with `turbidity` given `colour.number`, and is correlated with
`colour.number` given `turbidity`. This suggests that, indeed, `secchi` depends
on both turbidity and water colour. (Using Spearman correlation as before.)

On the other hand, `turbidity` appears to be correlated with `secchi` given
colour but not with `colour.number` given `secchi`. The marginal correlation
between `turbidity` and `colour.number` does not appear to be particularly strong,
either. Therefore, it is likely that `turbidity` will not be linked to
`water.colour` even if we drop `secchi`.

```{r relationship_secchi_turbidity_colour, echo = FALSE}
local({

  df = data[, c("secchi", "turbidity", "colour.number")]    
  cormat = cor(df, use = "pairwise.complete", method = "spearman")
  precision = cov2cor(MASS::ginv(cormat))
  dimnames(precision) = dimnames(cormat)
  p1 = standard.heatmap(cormat, abs = TRUE)
  p2 = standard.heatmap(precision, abs = TRUE)
  grid.arrange(p1, p2, ncol = 2)

})
```

---

Water salinity seems to increase as you move from north-east to south-west, away
from Finland and towards Denmark and the open ocean.

```{r salinity_by_location, echo = FALSE, fig.width = 4, fig.height = 6}
local({

  props = prepare.map(data,
            fun = function(loc) mean(loc[, "salinity"], na.rm = TRUE))
  standard.heatmap(props)

})
```

The distribution of salinity does not appear to change over time.

```{r salinity_over_time, echo = FALSE, fig.width = 16, fig.height = 5}
local({
  
  figures = vector(nlevels(decades), mode = "list")
  names(figures) = levels(decades)

  for (d in levels(decades)) {

    props = prepare.map(data[decades == d, ],
              fun = function(loc) mean(loc[, "salinity"], na.rm = TRUE))
    figures[[d]] = standard.heatmap(props, numbers = FALSE, main = d)

  }#FOR

  do.call(grid.arrange, c(figures, list(ncol = 5)))

})
```

Salinity does not present any particular seasonal patterns.

```{r salinity_by_season, echo = FALSE, fig.width = 8, fig.height = 12}
local({

  figures = vector(nlevels(seasons), mode = "list")
  names(figures) = levels(seasons)

  for (s in levels(seasons)) {

    props = prepare.map(data[seasons == s, ],
              fun = function(loc) mean(loc[, "salinity"], na.rm = TRUE))
    figures[[s]] = standard.heatmap(props, main = s, cex = 0.6)

  }#FOR

  do.call(grid.arrange, c(figures, ncol = 2))

})
```


```{r}
str(data)
```
